Financialisation of Commodity Markets

### 2025-07-16 blind.tex
Deploying to gh-pages from @ bautheac/financialisation-co-movement@83fc75059bf36178bab2418157f7606c553b8174 🚀

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Financialisation of Commodity Markets},
  pdfauthor={Devraj Basu; Olivier Bauthéac},
  pdfkeywords={Commodity markets, Financialisation, Co-movement, Asset
pricing},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Financialisation of Commodity Markets \\\large{Co-movement
Behind-the-Scenes} }
\author[1]{Devraj Basu%
\corref{cor1}%
}
 \ead{devraj.basu@strath.ac.uk} 
\author[1]{Olivier Bauthéac%
%
}
 \ead{olivier.bautheac@strath.ac.uk}

\affiliation[1]{organization={University of Strathclyde Business
School, Accounting \& Finance},addressline={199 Cathedral
Street},city={Glasgow},postcode={G4 0QU},postcodesep={}}

@@

The returns pull-back continues into the post-crisis period,
interestingly amidst a lower volatility environment, suggesting that
investment outflows from the commodity futures complex have had a lower
impact on volatility than the corresponding inflows witnessed during
financialisation. The pattern of reversion toward the
pre-financialisation Keynesian paradigm continues stronger with mean
returns higher during phases of \textbf{backwardation} relative to
\textbf{contango} and the difference more pronounced than over the
crisis period. In contrast to the crisis period when they seemed to be
converging, the pricing performance of CHP and \textbf{CHP} now part
ways, seemingly reverting to the Keynesian paradigm observed
pre-financialisation, further suggesting a reversion to legacy commodity
specific dynamics. The pull back extends to the correlation structure
with an across-the-board decrease in average pairwise correlations down
to levels below those observed during financialisation. The average
market factor beta for the entire cross-section shows a similar pattern,
dropping down below the level observed during financialization with the
decline seemingly driven by the metals sector, particularly the GB
traded segment. The pattern extends to the pricing performance of all
factors that drops below levels observed during financialisation.\\
Moreover, both the average correlations and the pricing performance of
the market factor are considerably stronger during \textbf{contango}
phases, unlike previous periods, which showed no substantial difference
between regimes with the pattern particularly pronounced for metal
assets globally, both US and GB traded. These results document the
emerging role/dominance of the \textbf{contango} regime post-crisis and
could provide a valuable basis for further research.

@@

pattern of increase in average return during financialisation with an
equally weighted portfolio of the US traded assets showing a mean
returns of 15.9\% significant at the 5\% level vs.~6.9\% with a 10\%
significance in the pre-financialisation period. The increase is even
more pronounced for an equally weighted portfolio of GB traded assets
(metals), 2.47\% (not significant) to 21.8\% (significant at the 10\%
level), which sets out the global nature of the phenomenon. Most
commodities show a higher mean return while five---Lumber (XCME),
Natural gas (XNYM), Palladium (XNYM), Platinum (XNYM) and Nickel-primary
(XLME)---show a lower figure with five of them significant---Crude
oil-WTI (XNYM), Heating oil (XNYM), Copper (XCEC), Copper (XLME) and
Tin-refined (XLME)---against none pre-financialisation. Sugar-\#11 wins
the race with a mean return increasing from -4.6\% to 24.9\% followed by
copper with sizeable increases for both markets: 1.5\% to 28.2\% (XCEC)
and 0.5\% to 28.6\% (XLME).\\
The pattern of results for volatilities is similar although the increase
is less pronounced. Both equally weighted portfolios shows an increase,
more pronounced for the GB assets, 10\% to 13.6\% (US) vs.~15.7\% to
26.9\% (GB), while out of the thirty assets, eleven show a drop in
volatility: Oats (XCBT), Lean hogs (XCME), Cocoa (IFUS), Coffee-C
(IFUS), Lumber (XCME), Sugar-\#11 (IFUS), Natural gas (XNYM), Crude
oil-WTI (XNYM), Gasoline (XNYM), Heating oil (XNYM), Palladium (XNYM).
Financialisation seems however to have impacted the commodity complex
beyond these mechanical effects of increased long participation ensuing
from the entry on index type investments as our results suggest next.\\

@@

\caption{\label{tbl-stats-regime-tests}This table shows mean returns and
volatilities for the individual commodities as well as for two equally
weighted portfolios of US and GB traded assets across the four periods.
For each period, estimates are computed independently over the whole
period (figure shown) as well as over phases of \textbf{backwardation}
(b) and \textbf{contango} (c) with corresponding comparison results
reported in brackets. The table further shows the results of t-tests of
significance for the mean returns computed over the whole periods as
well as t-tests of difference between these when computed over the two
phases independently along with F-tests of difference for the
corresponding variances. Values significant at the 1\%, 5\% and 10\%
level are marked with ***, ** and * respectively. The estimate values
per regime as well as the p-values for the corresponding difference
tests are available in the attached
\href{http://bautheac.duckdns.org:3838/bautheac/co-movement/}{Internet
Appendix}. \textbf{Backwardation} and \textbf{contango} regimes as well
as period definitions are discussed in Section~\ref{sec-methods} while
the results are discussed in Section~\ref{sec-results}.}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Financialisation of Commodity Markets},
  pdfkeywords={Commodity markets, Financialisation, Co-movement, Asset
pricing},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Financialisation of Commodity Markets \\\large{Co-movement
Behind-the-Scenes} }

@@

The returns pull-back continues into the post-crisis period,
interestingly amidst a lower volatility environment, suggesting that the
underlying contributing factors--including investment outflows from the
commodity futures complex as well as perhaps the evolving nature of
global oil demand over the period--have had a lower impact on volatility
than witnessed during financialisation. The pattern of reversion toward
the pre-financialisation Keynesian paradigm continues stronger with mean
returns higher during phases of \textbf{backwardation} relative to
\textbf{contango} and the difference more pronounced than over the
crisis period. In contrast to the crisis period when they seemed to be
converging, the pricing performance of CHP and \textbf{CHP} now part
ways, seemingly reverting to the Keynesian paradigm observed
pre-financialisation, further suggesting a reversion to legacy commodity
specific dynamics. The pull back extends to the correlation structure
with an across-the-board decrease in average pairwise correlations down
to levels below those observed during financialisation. The average
market factor beta for the entire cross-section shows a similar pattern,
dropping down below the level observed during financialization with the
decline seemingly driven by the metals sector, particularly the GB
traded segment. The pattern extends to the pricing performance of all
factors that drops below levels observed during financialisation.\\
Moreover, both the average correlations and the pricing performance of
the market factor are considerably stronger during \textbf{contango}
phases, unlike previous periods, which showed no substantial difference
between regimes with the pattern particularly pronounced for metal
assets globally, both US and GB traded. These results document the
emerging role/dominance of the \textbf{contango} regime post-crisis and
could provide a valuable basis for further research.

@@

pattern of increase in average return during financialisation with most
individual commodities showing a higher mean return and an equally
weighted portfolio of the US traded assets showing a mean returns of
15.9\% significant at the 5\% level vs.~6.9\% with a 10\% significance
in the pre-financialisation period. The increase is even more pronounced
for an equally weighted portfolio of GB traded assets (metals), 2.47\%
(not significant) to 21.8\% (significant at the 10\% level), which sets
out the global nature of the phenomenon. The pattern of results for
volatilities is similar although the increase is less pronounced with a
weaker majority of individual commodities showing heightened volatility
and both equally weighted portfolios showing an increase, more
pronounced for the GB assets. Financialisation seems however to have
impacted the commodity complex beyond these mechanical effects of
increased long participation ensuing from the entry of index type
investments as our results suggest next.\\

@@

\caption{\label{tbl-stats-regime-tests}This table shows mean returns and
volatilities for the individual commodities and two equally weighted
portfolios of US and GB traded assets for the four periods. For each
period, estimates are computed independently over the whole
period--figure shown--as well as over phases of \textbf{backwardation}
(b) and \textbf{contango} (c)--results in brackets. The table further
shows the results of t-tests of significance for the mean returns
computed over the whole periods and t-tests of difference between these
when computed over the two phases independently along with F-tests of
difference for the corresponding variances. Values significant at the
1\%, 5\% and 10\% level are marked with ***, ** and * respectively.
Detailed results are available in the attached
\href{http://bautheac.duckdns.org:3838/bautheac/co-movement/}{Internet
Appendix}. \textbf{Backwardation} and \textbf{contango} regimes as well
as period definitions are discussed in Section~\ref{sec-methods} while
the results are discussed in Section~\ref{sec-results}.}

### 2025-07-16 paper.qmd
Deploying to gh-pages from @ bautheac/financialisation-co-movement@83fc75059bf36178bab2418157f7606c553b8174 🚀

---
title: Financialisation of Commodity Markets
subtitle: Co-movement Behind-the-Scenes
author:
  - name: Devraj Basu
    email: devraj.basu@strath.ac.uk
    affiliations: 
        - id: strathclyde
          name: University of Strathclyde Business School
          department: Accounting & Finance
          address: 199 Cathedral Street
          city: Glasgow
          state: Scotland
          postal-code: G4 0QU
    attributes:
        corresponding: true
  - name: Olivier Bauthéac
    email: olivier.bautheac@strath.ac.uk
    affiliations:
        - ref: strathclyde
abstract: |
  In the early 2000s institutional investors entered the commodity futures markets en masse with passive, long only, index type positions in sharp contrast with those typically assumed by traditional expert participants. A heated public debate soon erupted over the perceived consequences of the phenomenon---commonly referred to as "financialisation"---and, in response to immediate policy concerns, the matter was thrust onto the academic sphere as a burning issue. With the benefit of hindsight it now seems that the academic debate was framed rather narrowly, used contentious research methods, and eventually led to regulatory changes that were therefore perhaps unwarranted. In contrast, we take a broader approach where we consider a large cross-section of liquid commodities as suggested by the nature of financialisation that comprehends commodity futures as an asset class. We examine the cross-sectional interconnectedness of these assets and uses a bespoke asset pricing factors based framework to study the issue through the lens of co-movement. We find that the phenomenon had ontological consequences for the commodity complex with its impact extending beyond the mechanical effects induced by indexation. The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, seem to have set off a motion of reversion to legacy pre-financialisation fundamentals.
keywords: 
  - Commodity markets
  - Financialisation
  - Co-movement
  - Asset pricing
date: last-modified
bibliography: bibliography.bib
code_repository: https://github.com/bautheac/financialisation-co-movement
format:
  elsevier-pdf:
    keep-tex: true
    journal:  
      name: Journal of Futures Markets
      formatting: preprint
      model: 3p
      cite-style: authoryear

  html:
    grid:
      body-width: 1000px
    toc: true
    toc_depth: 3
    toc-title: Contents
    css: style.css
    code-links:
      - text: github repository
        icon: github
        href: https://github.com/bautheac/financialisation-co-movement
---

@@

@tbl-stats-combined displays descriptive statistics for the commodity series considered in the study; the results show a clear pattern of increase in average return during financialisation with an equally weighted portfolio of the US traded assets showing a mean returns of 15.9% significant at the 5% level vs. 6.9% with a 10% significance in the pre-financialisation period. The increase is even more pronounced for an equally weighted portfolio of GB traded assets (metals), 2.47% (not significant) to 21.8% (significant at the 10% level), which sets out the global nature of the phenomenon. Most commodities show a higher mean return while five---Lumber (XCME), Natural gas (XNYM), Palladium (XNYM), Platinum (XNYM) and Nickel-primary (XLME)---show a lower figure with five og them significant---Crude oil-WTI (XNYM), Heating oil (XNYM), Copper (XCEC), Copper (XLME) and Tin-refined (XLME)---against none pre-financialisation. Sugar-#11 wins the race with a mean return increasing from -4.6% to 24.9% followed by copper with sizeable increases for both markets: 1.5% to 28.2% (XCEC) and 0.5% to 28.6% (XLME).  
The pattern of results for volatilities is similar although the increase is less pronounced. Both equally weighted portfolios shows an increase, more pronounced for the GB assets, 10% to 13.6% (US) vs. 15.7% to 26.9% (GB), while out of the thirty assets, eleven show a drop in volatility: Oats (XCBT), Lean hogs (XCME), Cocoa (IFUS), Coffee-C (IFUS), Lumber (XCME), Sugar-#11 (IFUS), Natural gas (XNYM), Crude oil-WTI (XNYM), Gasoline (XNYM), Heating oil (XNYM), Palladium (XNYM). Financialisation seems however to have impacted the commodity complex beyond these mechanical effects of increased long participation ensuing from the entry on index type investments as our results suggest next.   
The cross-sectional nature of financialisation naturally leads us to extend the individual hedging pressure paradigm by incorporating aggregate CHP (**CHP**), a market wide measure of hedging pressure. The traditional Keynesian hedging pressure paradigm postulates a negative relationship between the return on an individual commodity futures and its hedging pressure. The extended Keynesian hedging pressure paradigm, in the context of financialisation, should also imply that returns for individual commodities should be high in periods where **CHP** is low and vice et versa. @tbl-stats-combined further shows the mean returns and volatilities for high (**contango**) and low (**backwardation**) **CHP** regimes for each time period while @tbl-regime-diference-tests shows the results of difference tests for returns mean and variance between the two regimes.  

@@

Financialisation was a global phenomenon, through the channel of index investment it simultaneously influenced the entire spectrum of liquid commodities, eventually spurring the transformation of commodity markets from a collection of idiosyncratic markets into a more cohesive collection akin to an asset class. In this context, interconnectedness emerges as a crucial aspect that warrants thorough examination.   
We strive to address this issue by first examining the average returns pairwise correlations between commodities across the entire cross-section of the commodity assets considered in the study as well as by country, sector and sub-sector. @tbl-correlations-inner-periods shows the results for the four periods of interest with the results reported for the entire period as well as for periods of **backwardation** and **contango**. For the entire sample the average pairwise correlation jumps from 0.07 in the first period to 0.16 over financialisation. This finding indicates higher interdependence within the full cross-section of commodities starting with the financialisation period, consistent with the empirical findings in @fryMcKibbin_evolution_2023 and @mayer_financialization_2017 as well as with the theoretical model of @basak_model_2016. The sector analysis sheds light on the driving force behind these changes. The US metals see the sharpest increase in pairwise correlations going from 0.22 over the past period to 0.56 over financialisation while the three other sectors see a much smaller percentage increase during financialisation relative to the past period. These results thus suggest that the US metals sector was the driving force behind increased average pairwise correlations during financialisation, a new observation to the best of our knowledge. The pattern of correlations is very similar across phases of **backwardation** and **contango** for all the sectors and sub-sectors over the past and financialisation periods. In this context, the ontological consequences of financialisation are better observed from an investment perspective as illustrated in @tbl-correlations-cross-periods where we observe average pairwise correlation coefficients between equally weighted portfolios of US traded commodity assets grouped by sub-sector. Although modest in magnitude, these exhibit a telling pattern, mirroring that shown by the individual commodity returns in @tbl-stats-combined, with a substantially higher average coefficient over **backwardation** phases in the first period (0.12 vs. 0.03) and the pattern reversed over financilisation with an average coefficient substantially higher in **contango** (0.12 vs. 0.02).  
We break down @tbl-correlations-inner-periods's average correlations by year in @tbl-correlations-inner-years. For the full sample of assets, the average pairwise correlations remain in the 0.06–0.08 range from 1998 through 2003, before suddenly increasing to 0.12 from 2003 to 2004. This suggests 2004 as a distinct break point for the onset of financialisation, as noted in several studies [@baker_financialization_2021; @tang_index_2012]. We further observe that the US metals appear to have played a major role in this as the increase seems driven by the sector which average pairwise correlation more than doubles (0.49 vs 0.22) over the period.

@@

@@

# Tables

```{r}
#| label: tbl-stats-combined
#| tbl-cap: "This table shows mean returns and corresponding volatility for the twenty four individual US commodities and the six GB metals considered in the study as well as for two equally weighted portfolios formed from the US commodities and the GB metals respectively across the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). for each period the figures are shown independently for the entire period as well as for phases of aggregate backwardation (aggregate CHP $\\leq$ period median) and aggregate contango (aggregate CHP > period median). Mean values significant at the 1%, 5% and 10% level are marked with \\\\*\\\\*\\\\*, \\\\*\\\\* and \\\\* respectively. Aggregate CHP construction and corresponding regime definitions are discussed in \\@sec-methods while the results are discussed in \\@sec-results."

table_name <- "stats - combined"
columns_to_collapse_rows_for <- c("asset", "regime")
colnames <- c("asset", "regime", "estimate", periods)
align <- make_series_of_repeated_chars(list(l = 3L, r = 4L))
longtable <- TRUE
bold_columns <- 1L:3L
coloured_columns <- 4L:7L

@@

```{r}
#| eval: false
#| label: tbl-experimental
#| tbl-cap: "lorem ipsum dolor sit amet consectetur adipiscing elit sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident sunt in culpa qui officia deserunt mollit anim id est laborum. Lorem ipsum dolor sit amet consectetur adipiscing elit sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident sunt in culpa qui officia deserunt mollit anim id est laborum."
#| tbl-subcap: 
#|     - "Average pairwise inner correlations"
#|     - "Average pairwise cross correlations"
#| layout-nrows: 2

table1_format <- list(
  table_name = "correlations - inner - periods",
  columns_to_collapse_rows_for = c("country", "sector", "subsector"),
  longtable = FALSE,
  colnames = c(sectors, "regime", periods),
  align = make_series_of_repeated_chars(list(l = 3L, r = 5L)),
  bold_columns = 1L:8L, 
  coloured_columns = 5L:8L
  )
table2_format <- list(
  table_name = "correlations - cross - periods",
  columns_to_collapse_rows_for = c("portfolios", "regime"),
  longtable = TRUE,
  colnames = c("portfolios", "regime", periods),
  align = make_series_of_repeated_chars(list(l = 1L, r = 5L)),
  bold_columns = 1L:2L, 
  coloured_columns = 3L:6L
)
formats <- list(table1_format, table2_format)

get_results(tables, table1_format$table_name) %>% dplyr::mutate(dplyr::across(dplyr::where(is.factor), as.character)) %>%
    collapse_rows(table1_format$columns_to_collapse_rows_for) %>%
    kable(booktabs = T, longtable = table1_format$longtable, col.names = table1_format$colnames, align = table1_format$align, escape = FALSE) %>%
    kable_styling(font_size = 7L) %>%
    row_spec(0L, bold = TRUE, color = "black") %>%
    column_spec(table1_format$bold_columns, bold = TRUE) %>% column_spec(table1_format$coloured_columns, color = "#4285f4")

get_results(tables, table2_format$table_name) %>% dplyr::mutate(dplyr::across(dplyr::where(is.factor), as.character)) %>%
    collapse_rows(table2_format$columns_to_collapse_rows_for) %>%
    kable(booktabs = T, longtable = table2_format$longtable, col.names = table2_format$colnames, align = table2_format$align, escape = FALSE) %>%
    kable_styling(font_size = 7L) %>%
    row_spec(0L, bold = TRUE, color = "black") %>%
    column_spec(table2_format$bold_columns, bold = TRUE) %>% column_spec(table2_format$coloured_columns, color = "#4285f4")
```

```{r}
#| label: tbl-correlations-inner-years
#| tbl-cap: "While \\@tbl-correlations-inner-periods shows results calculated independently for the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018), this stable shows the results of the same analysis calculated by year over the entire sample period (1997-2018). The results are discussed in \\@sec-results while \\@sec-taxonomy details the taxonomy for the commodity assets considered."

>>>>

---
title: Financialisation of Commodity Markets
subtitle: Co-movement Behind-the-Scenes
author:
  - name: Devraj Basu
    email: devraj.basu@strath.ac.uk
    affiliations: 
        - id: strathclyde
          name: University of Strathclyde Business School
          department: Accounting & Finance
          address: 199 Cathedral Street
          city: Glasgow
          state: Scotland
          postal-code: G4 0QU
    attributes:
        corresponding: true
  - name: Olivier Bauthéac
    email: olivier.bautheac@strath.ac.uk
    affiliations:
        - ref: strathclyde
abstract: |
  In the early 2000s institutional investors entered the commodity futures markets en masse with passive, long only, index type positions in sharp contrast with those typically assumed by traditional expert participants. A heated public debate soon erupted over the perceived consequences of the phenomenon---commonly referred to as "financialisation"---and, in response to immediate policy concerns, the matter was thrust onto the academic sphere as a burning issue. With the benefit of hindsight it now seems that the academic debate was framed rather narrowly, used contentious research methods, and eventually led to regulatory changes that were therefore perhaps unwarranted. In contrast, we take a broader approach where we consider a large cross-section of liquid commodities as suggested by the nature of financialisation that comprehends commodity futures as an asset class. We examine the cross-sectional interconnectedness of these assets and uses a bespoke asset pricing factors based framework to study the issue through the lens of co-movement. We find that the phenomenon had ontological consequences for the commodity complex with its impact extending beyond the mechanical effects induced by indexation. The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, seem to have set off a motion of reversion to legacy pre-financialisation fundamentals.
keywords: 
  - Commodity markets
  - Financialisation
  - Co-movement
  - Asset pricing
date: last-modified
bibliography: bibliography.bib
code_repository: https://github.com/bautheac/financialisation-co-movement
format:
  elsevier-pdf:
    keep-tex: true
    journal:  
      name: the Journal of Futures Markets
      formatting: preprint
      model: 3p
      cite-style: authoryear
  html:
    grid:
      body-width: 1000px
    toc: true
    toc_depth: 3
    toc-title: Contents
    css: style.css
    code-links:
      - text: github repository
        icon: github
        href: https://github.com/bautheac/financialisation-co-movement
---

@@

@tbl-stats-regime-tests displays descriptive statistics for the commodity series considered in the study; the results show a clear pattern of increase in average return during financialisation with most individual commodities showing a higher mean return and an equally weighted portfolio of the US traded assets showing a mean returns of 15.9% significant at the 5% level vs. 6.9% with a 10% significance in the pre-financialisation period. The increase is even more pronounced for an equally weighted portfolio of GB traded assets (metals), 2.47% (not significant) to 21.8% (significant at the 10% level), which sets out the global nature of the phenomenon. The pattern of results for volatilities is similar although the increase is less pronounced with a weaker majority of individual commodities showing heightened volatility and both equally weighted portfolios showing an increase, more pronounced for the GB assets. Financialisation seems however to have impacted the commodity complex beyond these mechanical effects of increased long participation ensuing from the entry of index type investments as our results suggest next.     
The cross-sectional nature of financialisation naturally leads us to extend the individual hedging pressure paradigm by incorporating aggregate CHP (**CHP**), a market wide measure of hedging pressure. The traditional Keynesian hedging pressure paradigm postulates a negative relationship between the return on an individual commodity futures and its hedging pressure. The extended Keynesian hedging pressure paradigm, in the context of financialisation, should also imply that returns for individual commodities should be high in periods where **CHP** is low and vice et versa. @tbl-stats-regime-tests further reports the dominating regime for each asset-period-moment combination along with the corresponding results of difference tests for returns mean and variance between the two regimes^[The estimate values per regime as well as the p-values for the corresponding difference tests are available in the attached [Internet Appendix](http://bautheac.duckdns.org:3838/bautheac/co-movement/).].  

@@

Financialisation was a global phenomenon, through the channel of index investment it simultaneously influenced the entire spectrum of liquid commodities, eventually spurring the transformation of commodity markets from a collection of idiosyncratic markets into a more cohesive collection akin to an asset class. In this context, interconnectedness emerges as a crucial aspect that warrants thorough examination.   
We strive to address this issue by first examining the average returns pairwise correlations between commodities across the entire cross-section of the commodity assets considered in the study as well as by country, sector and sub-sector. @tbl-correlations-inner-periods shows the results for the four periods of interest with the results reported for the entire period as well as for periods of **backwardation** and **contango**. For the entire sample the average pairwise correlation jumps from 0.07 in the first period to 0.16 over financialisation. This finding indicates higher interdependence within the full cross-section of commodities starting with the financialisation period, consistent with the empirical findings in @fryMcKibbin_evolution_2023 and @mayer_financialization_2017 as well as with the theoretical model of @basak_model_2016. The sector analysis sheds light on the driving force behind these changes. The US metals see the sharpest increase in pairwise correlations going from 0.22 over the past period to 0.56 over financialisation while the three other sectors see a much smaller percentage increase during financialisation relative to the past period. These results thus suggest that the US metals sector was the driving force behind increased average pairwise correlations during financialisation, a new observation to the best of our knowledge. The pattern of correlations is very similar across phases of **backwardation** and **contango** for all the sectors and sub-sectors over the past and financialisation periods. In this context, the ontological consequences of financialisation are better observed from an investment perspective as illustrated in @tbl-correlations-cross-periods where we observe average pairwise correlation coefficients between equally weighted portfolios of US traded commodity assets grouped by sub-sector^[A more detailed breakdown of country, sector and subsector average pairwise correlations is provided in the attached [Internet Appendix](http://bautheac.duckdns.org:3838/bautheac/co-movement/).]. Although modest in magnitude, these exhibit a telling pattern, mirroring that shown by the individual commodity returns in @tbl-stats-regime-tests, with a substantially higher average coefficient over **backwardation** phases in the first period (0.12 vs. 0.03) and the pattern reversed over financilisation with an average coefficient substantially higher in **contango** (0.12 vs. 0.02).  
We break down @tbl-correlations-inner-periods's average correlations by year in @tbl-correlations-inner-years. For the full sample of assets, the average pairwise correlations remain in the 0.06–0.08 range from 1998 through 2003, before suddenly increasing to 0.12 from 2003 to 2004. This suggests 2004 as a distinct break point for the onset of financialisation, as noted in several studies [@baker_financialization_2021; @tang_index_2012]. We further observe that the US metals appear to have played a major role in this as the increase seems driven by the sector which average pairwise correlation more than doubles (0.49 vs 0.22) over the period.

@@

# Tables {.unnumbered}

@@

```{r}
#| label: tbl-stats-regime-tests
#| tbl-cap: "This table shows mean returns and volatilities for the individual commodities and two equally weighted portfolios of US and GB traded assets for the four periods. For each period, estimates are computed independently over the whole period--figure shown--as well as over phases of **backwardation** (b) and **contango** (c)--results in brackets. The table further shows the results of t-tests of significance for the mean returns computed over the whole periods and t-tests of difference between these when computed over the two phases independently along with F-tests of difference for the corresponding variances. Values significant at the 1%, 5% and 10% level are marked with \\\\*\\\\*\\\\*, \\\\*\\\\* and \\\\* respectively. Detailed results are available in the attached [Internet Appendix](http://bautheac.duckdns.org:3838/bautheac/co-movement/). **Backwardation** and **contango** regimes as well as period definitions are discussed in \\@sec-methods while the results are discussed in \\@sec-results."

table_name <- "stats-individuals - regime difference tests"
columns_to_collapse_rows_for <- c("asset")
colnames <- c("asset", "estimate", "past", "financialisation", "crisis", "post-crisis")
align <- make_series_of_repeated_chars(list(l = 1L, r = 5L))
longtable <- TRUE
bold_columns <- 1L:6L 
coloured_columns <- 3L:6L

@@

```{r}
#| label: tbl-correlations-inner-years
#| tbl-cap: "While \\@tbl-correlations-inner-periods shows results computed independently for the four periods, this table shows the results of the same analysis implemented for each year over the entire sample period. The results are discussed in \\@sec-results while \\@sec-taxonomy details the commodity assets taxonomy."
